Headgroup structure and cation binding in phosphatidylserine lipid bilayers

### 2019-08-15 Manuscript/manuscriptPS.tex
Updates based on reviewer's comments

\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid type in eukaryotic biomembranes. In red blood cells, for example,
PS lipids compose 8.5\% of the total lipid weight. 
The abundance, however, varies between different organelles, and up to
25--35\% of the cytosolic leaflet of plasma membranes 
\cite{lemmon08,leventis10,li14} consists of PS lipids.
PS lipids are vastly important biomolecules that interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some protein domains interact specifically with PS lipids,
while other protein sites attract PS lipids by nonspecific electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, deciphering the structural details
of lipid headgroups and the details of cation binding
is crucial for understanding PS-mediated processes on cellular membranes.

@@

The molecular level interpretation of these observations is currently lacking. Classical molecular dynamics (MD) simulations have been widely used in efforts 
to understand the PS headgroup structure, its influence on lipid bilayer properties, and its
interaction with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
Unfortunately, the results have depended on the particular force field used.
For example, recent simulations using the NBfix parameters for Ca$^{2+}$~\cite{kim16} in
the CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}. In contrast,
the results from the same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
rotational-echo double-resonance (REDOR) experiments, indicate a significant binding affinity also toward the phosphate region \cite{hallock18}.
Meanwhile, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest substantial
calcium binding to the carbonyls in the acyl chains \cite{melcrova16}.

@@

The $S_{\rm CH}$ of the $\alpha$ and $\beta$ carbons in the PC headgroup
decrease proportionally to the amount of positive
charge bound to the bilayer \cite{akutsu81,altenbach84,seelig87},
and can therefore be used to measure the ion binding affinity.
This concept, known as the molecular electrometer, is especially useful for 
comparison between simulations and experiments, as
the headgroup $S_{\rm CH}$ at varying cation
concentrations can be easily calculated from
simulations~\cite{catte16}. The headgroup $S_{\rm CH}$
of negatively charged PS and PG lipids also exhibit systematic --- although less well understood ---dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, measuring the PC headgroup $S_{\rm CH}$ from 
mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also SI section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing the ion binding to negatively charged membranes.

Calibrating the PC $S_{\rm CH}$ response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer.
This can be done using experimental data from mixtures of PC and
monovalent cationic surfactants (such as POPC and dihexadecyldimethylammonium,
see SI section \ref{electrometerCALIBRATION})~\cite{scherer89,melcr18}.
Additionally, we here quantify the response of PC headgroup $S_{\rm CH}$
to the negatively charged PS (see SI section \ref{electrometerFORmixtures}), which also follows the molecular electrometer
in experiments~\cite{scherer87}.

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
    Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}.
    Here {\textsf{\small M}} indicates a magnitude problem, {\textsf{\small F}} a forking problem; letter size increases with problem severity. Color scheme: "within experimental error" (dark green), "almost within experimental error" (light green), "clear deviation from experiments" (light red), and "major deviation from experiments" (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see Supplementary Information.
  }
\end{figure}

@@

In simulations that have the best agreement with experiments (CHARMM36 and Slipids),
the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
angle distribution for PS than for PC headgroup (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
These results might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}, and the suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments. However, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to elucidate the correct conformational ensemble.

@@

First, we complemented the available experimental information \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters,
and then proceeded to compare MD simulation results from several force fields with the experimental data.
This revealed that none of the force fields
reproduce the PS headgroup order parameters within the experimental accuracy. However,
the best models for the PS headgroup suggested a characteristic rigid conformation for its
carboxyl group. Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
concentrations \cite{roux90} then showed that the tested MD force fields
overestimate the cation binding affinity to these bilayers with two exceptions: 1) the MacRog simulation with potassium counterions appears to produce the most realistic monovalent ion binding
affinity to PS-containing lipid bilayers, and 2) the CHARMM36 force field with the recently introduced
NBfix correction for calcium \cite{kim16} underestimates the calcium binding affinity.
The experimentally measured response of the PS headgroup order parameters to the bound calcium, as well as to the dilution of bilayer with zwitterionic PC lipids, were not
qualitatively reproduced in any of the tested force fields. This underlines the need for more accurate
MD force fields to study the biological function of PS lipids.
This is in contrast to the previous results with PC lipids,
where the experimentally measured headgroup order parameter responses to the bound charge
were in qualitative agreement even though the headgroup structures themselves were
incorrect and the cation binding affinities were overestimated \cite{catte16}.

We expect our results to pave the way for the development of better MD force fields
for PS lipids. The quality of conformational ensembles produced by the various models can be evaluated based on the headgroup order parameters, which we hope to guide the development of force fields towards models that correctly reproduce the PS headgroup structures. The cation binding can be improved based on the experimental headgroup order parameter data from POPC:POPS (5:1) mixtures under different cation concentrations. This was already demonstrated for POPC using the electronic continuum correction \cite{melcr18}, and a similar study for POPS is being progressed separately \cite{ECCpops}.

>>>>

\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid type in eukaryotic biomembranes. In red blood cells, for example,
PS lipids compose 8.5\% of the total lipid weight~\cite{lemmon08}. 
The abundance, however, varies between different cells and organelles, and up to
25--35\% of the cytosolic leaflet of plasma membranes consists of PS lipids~\cite{leventis10,li14}.
PS lipids are vastly important biomolecules that interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some protein domains interact specifically with PS lipids,
while other protein sites attract PS lipids by nonspecific electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, deciphering the structural details
of lipid headgroups and the details of cation binding
is crucial for understanding PS-mediated processes on cellular membranes.

@@

Molecular level interpretations of the rigidity of PS headgroup and its interactions with ions
are currently lacking. Classical molecular dynamics (MD) simulations have been widely used in efforts 
to understand the PS headgroup structure, its influence on lipid bilayer properties, and its
interaction with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
Unfortunately, the results have depended on the particular force field used.
For example, recent simulations using the NBfix parameters for Ca$^{2+}$~\cite{kim16} in
the CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}. In contrast,
the results from the same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
rotational-echo double-resonance (REDOR) experiments, indicate a significant binding affinity also toward the phosphate region \cite{hallock18}.
Meanwhile, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest substantial
calcium binding to the carbonyls in the acyl chains \cite{melcrova16}.

@@

The $S_{\rm CH}$ of the $\alpha$ and $\beta$ carbons in the PC headgroup
decrease proportionally to the amount of positive
charge bound to the bilayer \cite{akutsu81,altenbach84,seelig87},
and can therefore be used to measure the ion binding affinity.
In addition to ions, the correlation between bound charge and headgroup order parameter change
is empirically observed also for peptides, charged amphiphiles, local anesthetics and charged lipids~\cite{scherer87,beschiaschvili91}.
This concept, known as the molecular electrometer, is especially useful for 
comparison between simulations and experiments, as
the headgroup $S_{\rm CH}$ at varying cation
concentrations can be easily calculated from
simulations~\cite{catte16}. The headgroup $S_{\rm CH}$
of negatively charged PS and PG lipids also exhibit systematic --- although less well understood ---dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, measuring the PC headgroup $S_{\rm CH}$ from 
mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also SI section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing the ion binding to negatively charged membranes.

Calibrating the PC $S_{\rm CH}$ response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer.
This can be done using experimental data from mixtures of PC and
monovalent cationic surfactants (such as POPC and dihexadecyldimethylammonium,
see SI section \ref{electrometerCALIBRATION})~\cite{scherer89,melcr18}.
Additionally, the response of PC headgroup $S_{\rm CH}$ to the negatively
charged PS follows the molecular electrometer in experiments~\cite{scherer87},
which we also quantify here (see SI section \ref{electrometerFORmixtures}).

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
    Rough subjective ranking of force fields based on how far are the simulated order parameters from the experimental
    values ({\textsf{\small M}} indicates a magnitude problem) and how well the difference between order parameters
    of the two hydrogens attached in the same carbon are reproduced ({\textsf{\small F}} indicates a forking problem)
    in Figure~\ref{HGorderParametersPS}.
    The letter size increases with problem severity. Color scheme: "within experimental error" (dark green), "almost within experimental error" (light green), "clear deviation from experiments" (light red), and "major deviation from experiments" (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see Supplementary Information.
  }
\end{figure}

@@

In simulations that have the best agreement with experiments (CHARMM36 and Slipids),
the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
angle distribution for PS than for PC headgroup (Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}),
which might reflect the increased rigidity proposed in the early experimental studies \cite{browning80,buldt81}.
The suggested characteristic conformations of the PS headgroup can be useful when interpreting experiments,
however, as the none of the tested models fully reproduces the experimental order parameters, more accurate MD force fields are required to elucidate the correct conformational ensemble.

@@

First, we complemented the available experimental information \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters,
and then proceeded to compare MD simulation results from several force fields with the experimental data.
This revealed that none of the force fields
reproduce the PS headgroup order parameters within the experimental accuracy.
However, the best models captured essential differences with PC and
suggested characteristic conformations of PS headgroups.
Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
concentrations \cite{roux90} then showed that the tested MD force fields
overestimate the cation binding affinity to these bilayers with two exceptions: 1) the MacRog simulation with potassium counterions appears to produce the most realistic monovalent ion binding
affinity to PS-containing lipid bilayers, and 2) the CHARMM36 force field with the recently introduced
NBfix correction for calcium \cite{kim16} underestimates the calcium binding affinity.
The experimentally measured response of the PS headgroup order parameters to the bound calcium, as well as to the dilution of bilayer with zwitterionic PC lipids, were not
qualitatively reproduced in any of the tested force fields. This underlines the need for more accurate
MD force fields to study the biological function of PS lipids.
This is in contrast to the previous results with PC lipids,
where the experimentally measured headgroup order parameter responses to the bound charge
were in qualitative agreement even though the headgroup structures themselves were
incorrect and the cation binding affinities were overestimated \cite{catte16}.

We expect our results to pave the way for the development of better MD force fields for PS lipids.
The quality of conformational ensembles produced by the various models can be evaluated
based on the headgroup order parameters, which we hope to guide the development of force fields towards
models that correctly reproduce the PS headgroup structures. Also, the ensembles observed in the simulations
with the most realistic headgroup conformations (CHARMM36 and Slipids) hint the direction for the force
field improvement. The cation binding can be improved based on the experimental headgroup order parameter
data from POPC:POPS (5:1) mixtures under different cation concentrations. Recent studies showed that the
cation binding to bilayers with POPC and POPS lipids can be improved by implicit inclusion of electronic
polarizability using the electronic continuum correction (ECC) \cite{melcr18,ECCpops}, suggesting that
the electronic polarizability effects are important for lipid-ion interactions but polarizable force fields
may not be necessary to correctly capture the ion binding to lipid bilayers.
